DATA 621 - Business Analytics and Data Mining

### 2020-10-28 HW3/HW3_Final_AA.Rmd
Update HW3_Final_AA.Rmd

This data set has complete cases, thus there is no need for imputation. Based on
some common summary statistics, there are more observations where the crime rate
is below the median. It is already apparent that some of the predictors varies
depending the crime rate. For instance, there is a noticeable difference in the
means of `age`, `lstat`, `rad`, and `zn` between the crime rate groups.

@@

## Data Distribution
For each predictors, we computed kernel density estimators to understand their
distribution. The following plots show how predictors are distributed between
areas where the crime rate is higher than the median (blue) and areas where the
crime rate is below the median (red). It is of interest to understand variables
that highlight large variations between the two groups.

@@

The density plots reveal that most of the data does not have a normal
distribution. There are some are heavily right tailed variables and others which
are multi-modal. The most Gaussian of the variables appears to be the one
related to the average number of rooms per dwelling: `rm`. Another
interesting predictor is `zn`, the proportion of residential land zoned for
large lots. Nearly 73% of the observations have a value of 0 with the remaining
spread widely between 1% and 100%. This may the source of possible distortion. A
third highly-skewed variable is `chas`, the indicator as to whether the lot
borders the Charles River. Out of the `r nobs` observations,
`r sum(DT$chas == 0)` do **not** border the river.

To provide another view, this time highlighting outliers, the data was analyzed
using boxplots.

@@

## Data Correlation
Below is a corrgram of the data. The shape of the ellipse reflects the strength
of the correlation and the color represents the direction. Thus, the diagonal
which represents the perfect correlation of each variable with itself is a blue
diagonal line. The shape and the color scheme reflect the guide at the bottom
of the image.

@@

Perhaps this is a reflection of the noticeable influence that socio-economic
status has on the crime rate of an area. It is known that crime is concentrated
in disadvantaged, urban neighborhoods in the United States. The economic
segregation suggests that affluent neighborhoods may be further away in terms of
distance, and as a result, disadvantaged areas are more attractive to crime
because the probability of success is higher even if the targets are not as
profitable. Consider areas with a higher socio-economics, these areas are
correlated have a lower crime rate. Moreover, industrial areas are correlated
with higher levels of nitrogen oxide concentrations. This may be an indication
of why these areas are less density with residents of a higher status. This
trend is also seen with the `ptratio` since a higher ratio means less funding
for public institutions, which is common in areas of lower status.

@@

# DATA PREPARATION
There are no missing values so none need to be imputed. Different preparation
techniques may be used for different models.

## Models 1 & 2
The variable `chas` will be removed due to having statistically insignificant
correlation with almost all the variables in the dataset. Next, outliers will be
tempered by capping from below at the 5th percentile and from above at the 95th
percentile.

@@

## Model 3
For model 3, very little data manipulation will be done. There may be valuable
information in the outliers and a reasonable model may be able to extract
something from `chas`. However, `chas` will be converted to a factor. Similarly,
`rad` is an index of accessibility, so it is almost certainly a factor, although
probably ordered. Given there are now two factors in the predictors, dummy
dummy variables need to be created. Lastly, the **labels** of `target` will be
changed from integer to character for compliance with some of R's naming
conventions. Based on the definitions, 0 will be mapped to *below* and 1 to
*above*.

# BUILD MODELS
First, we will split the training data into two sets. This will allow us to
perform a cross validation scheme on the models to tune parameters for optimum
performance.

@@

## Model 1
### Base Model
We will start with a simple logistic regression model to serve as a baseline.
This includes all variables in the dataset.

@@

### Enhanced Model
We will use **backwards stepwise regression** to remove features that are not
statistically significant in predicting the target. The result is a model that
includes the following features: `zn`, `nox`, `age`, `dis`, `rad`, `tax`,
`ptratio`, `lstat`, and `medv`.

### Coefficient Discussion
It is important to note that the coefficients from the model are predicting
whether or not the target variable is a **1** (the crime rate is *above* the
median value). Additionally, it must be noted that the numeric coefficients are
relative to the range of values that the variable encompasses. What this means
is that it's possible to have a coefficient that seems small when we look at the
absolute magnitude, but that actually has a very strong effect when applied to
the data.

@@

Let's take a look at the *signs* of the coefficients. We can see that the
coefficients for the variables `zn` and `tax` are negative. This is indicative
of an inverse relationship; i.e., the higher these values, the less likely the
crime rate is above the median. The relationship for the zone variable aligns
with the findings from our initial data exploration. However, the relationship
for the tax variable does not; in fact, we saw that there was a positive
correlation between the two (as the tax rate increases, so does the probability
that the crime rate will be above the median).

The rest of the variables have positive coefficients. Many of these are
expected. For example, in our exploratory data analysis, we noticed that the
`nox` variable (nitrogen oxides concentration (parts per 10 million)) has the
greatest positive correlation with the target variable. This is surprising; we
hadn't thought that the nitrogen oxide concentration would have as large of an
impact on the crime rate, but some quick googling shows us that there may
actually be a relationship
([https://www.sciencedaily.com/releases/2019/10/191003114007.htm](Source)).

@@

We again see that the primary predictor in the logistic regression is the `nox`
with `rad` and `ptratio` rounding out the top three predictors. This is in line
with the base model. Living close to the center of town, having teachers, and
not having polluted air are the largest predictors in terms of above median crime.

@@

Model 3 will be a binary logistic regression allowing for all variables and
selected pair-wise interactions pruned using both forward and backward stepwise
regression based on AICc to select the optimal parameters. The captured indicies
of the test set for models 1 & 2 will be used to split the data for model 3 as
well. With the vanilla logistic regression model in R, `glm` with a binomial
family and a logit link, there are no other hyperparameters available for
tuning. More sophisticated logistic regression implementation have these
parameters for fine-tuning. There is also no reasonable way to implement
cross-validation so none will be done.

To consider *all* interactions would be foolhardy. Given the need for dummy
variables, there will 19 predictors and so \(_{19}C_2 = 171\) possible pairs.
This is too much. Therefore, intentional selection of pairs is required.
Looking at the predictors, there are some which logically may interact with
each other, such as:

  * `indus` and `nox`
    * Nitrous Oxide is a common industrial pollutant. There is a high
    correlation in the training set between the two. Allowing for its
    interaction may temper the result of their relationship without losing more
    information

@@

It seems that there may be reason to consider an interaction between `chas` and
`nox`. Can it be that many industrial plants are on the river? Similarly with
`indus` and `lstat`. Higher-cost homes tend to be further away from the traffic
and pollution of industrial areas.

@@

The selected model has some interesting results. Firstly, the predictor with the
greatest effect is `nox`. Further, the only model variable which serves to
**decrease** the crime rate from the baseline is `rm`. Other variables, such as
`chas`

The interactions between `indus` and both `nox` and `lstat` are significant. As
theorized, these interactions serve to temper their individual contributions to
crime. Remember, a 1 indicates crime above the median, so positive coefficients
indicate contributions to *increasing* the crime rate. A similar observation
holds `rm`. While `dis` on its own did not decrease the model's AICc, its
interaction to temper `rm` did help the model's deviance. No other selected
interactions were left in the model.

A number of variables fell out of the model, such as `chas`, `dis`, and `zn`
among others. The remaining predictors are all significant, at least at the 10%
level, except for `rad.24`. However, from the last trace of the stepping routine
shown below, removing it would ***drastically*** reduce the models performance.
Lastly, while `rad.1` is the base case, `rad.2` and `rad.6` have disappeared as
well. This indicates that both `rad = 2` and `rad = 6` should be considered
indistinguishable from `rad = 1`. 
```
                    Df Deviance    AIC
<none>                   90.565 122.56
+ dis                1   89.191 123.19
+ zn                 1   89.558 123.56
+ tax                1   89.700 123.70
- ptratio            1   93.763 123.76
+ age                1   90.156 124.16
+ `nox:chasOnRiver`  1   90.224 124.22
- `rm:dis`           1   94.254 124.25
+ chas.OnRiver       1   90.273 124.27
+ rad.6              1   90.315 124.31
- lstat              1   94.364 124.36
+ `ptratio:medv`     1   90.494 124.49
+ rad.2              1   90.554 124.55
- rm                 1   95.001 125.00
- `indus:lstat`      1   95.633 125.63
- `indus:nox`        1   97.351 127.35
- indus              1   97.597 127.60
- medv               1   98.568 128.57
- rad.3              1  101.594 131.59
- rad.5              1  105.594 135.59
- rad.7              1  118.832 148.83
- nox                1  134.786 164.79
- rad.8              1  149.967 179.97
- rad.4              1  152.757 182.76
- rad.24             1  152.797 182.80
```

@@

## Performance 
Below are the confusion matrices for the three models, followed by a table with
the selected performance.

>>>>

We can make some initial observaations. The data set has complete cases, thus, 
there is no need for imputation. Based on some common summary statistics, 
there are more observations where the crime rate is below the median. It is already 
apparent that some of the predictors varies depending the crime rate. For instance, 
there is a noticeable difference in the means of `age`, `lstat`, `rad`, and `zn` between 
the crime rate groups.

@@

## Data Distribution

For each predictors, we computed and drew kernel density estimate to understand 
their distribution. Using density plots, we show how predictors are distributed 
between areas where crime rate is higher than the median, i.e. blue, and areas 
where crime rate is below the median, i.e. red. Our focus is to understand 
variables that highlight large variations between the two groups.

@@

The plots reveal that most of the data does not have a normal distribution. 
As expected, we see the heavily right tailed variables and others, which are 
multi-modal. The most Gaussian of the variables appears to be the one related 
to the average number of rooms per dwelling: `rm`.

Another interesting predictor, `zn` i.e. proportion of residential land zoned 
for large lots, has a significant positive skew (skew = 2.18). Nearly 73% of 
the observations (339 of 466 total) have a value of zero. When analyzing the 
difference between the crime rate groups, it is possible that areas with high 
crime rate do not have land zoned for large lots suggesting suburban areas are 
likely to have low crime rates because there typically have large lots, whereas 
urban areas have a higher crime rate since lot size are smaller than 25000 square 
feet.

Lastly, a third highly skewed variable is `chas`, the indicator as to whether 
the lot borders the Charles River. Out of the `r nobs` observations,
`r sum(DT$chas == 0)` do **not** border the river.

To provide another view, this time highlighting outliers, we present the data and 
analyze using boxplots.

@@

## Data Correlation

@@

It could be possible we are seeing a reflection of the noticeable influence 
that socio-economic status has on the crime rate of an area.  Graif, Gladfelter 
and Matthews (2016) state crime is generally concentrated in disadvantaged, 
urban neighborhoods in the United States.  An economic segregation suggests 
that affluent neighborhoods may be further away in terms of distance, and as 
a result, disadvantaged areas are more attractive to crime because the probability 
of success is higher even if the targets are not as profitable. We prove this data, 
showing areas with a higher socio-economics correlating to a lower crime rate.

Fields (2004) states industrial areas correlate with higher levels of nitrogen oxide 
concentrations. It may be an indication of why these areas are less dense with residents 
of a higher status. We see the same trend in `ptratio` since a higher ratio means less 
funding for public institutions, which is common in areas of lower status.

([https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4928692/](Source)).
([https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1247398//](Source)).

@@

# DATA PREPARATION

As there are no missing values, we have no need to impute. Instead, different preparation
techniques are utilized for different models.

## Models 1 & 2

The variable `chas` will be removed due to having statistically insignificant correlation 
with almost all the variables in the dataset.  Next, outliers are tempered by capping below 
the 5th percentile and above the 95th percentile.

@@

## Model 3

For model 3, we chose very little data manipulation, as there may be valuable 
information in the outliers.  A reasonable model may be able to extract 
something from `chas`.  Some variables, such as `chas` and `rad` (an index of 
accessibility) are converted to a factor. Additionally, `rad` probably ordered. 
Given there are now two factors in the predictors, we require creation of dummy 
variables. Lastly, the **labels** of `target` will be changed from integer to 
character for compliance with some of R's naming conventions.  Based on the 
definitions, 0 will be mapped to *below* and 1 to *above*.

# BUILD MODELS

First, we will split the training data into two sets. This will allow us to
perform a cross validation scheme on the models to tune parameters for optimum
performance.

@@

## Model 1
### Base Model

We will start with a simple logistic regression model to serve as a baseline.
This includes all variables in the dataset.

@@

### Enhanced Model

We will use **backwards stepwise regression** to remove features that are not
statistically significant in predicting the target. The result is a model that
includes the following features: `zn`, `nox`, `age`, `dis`, `rad`, `tax`,
`ptratio`, `lstat`, and `medv`.

### Coefficient Discussion

It is important to note that the coefficients from the model are predicting 
whether or not the target variable is a **1** (the crime rate *above* the 
median value). Additionally, since the numeric coefficients are relative to 
the range of values that the variable encompasses it is possible to have a 
coefficient that seems small when we look at the absolute magnitude, but that 
actually has a very strong effect when applied to the data.

@@

Looking at the *signs* of the coefficients, we can see the variables `zn` 
and `tax` are negative. This is indicative of an inverse relationship; 
i.e., the higher these values, the less likely the crime rate is above the 
median. The relationship for the zone variable aligns with the findings from 
our initial data exploration. However, the relationship for the tax variable 
does not; in fact, we saw that there was a positive correlation between the 
two variables.  As the tax rate increases so does the probability that the 
crime rate will be above the median.

The rest of the variables have positive coefficients and many of these are 
expected. For example, in our exploratory data analysis, we noticed that the 
`nox` variable (nitrogen oxides concentration (parts per 10 million)) has the 
greatest positive correlation with the target variable. This is surprising!  
We had not thought that the nitrogen oxide concentration would have as large 
of an impact on the crime rate. According to Colorado State University (2019), 
there may actually be a strong relationship between the two.  The team of 
researchers found strong correlation between exposure to air pollution and 
aggressive behavior.

([https://www.sciencedaily.com/releases/2019/10/191003114007.htm](Source)).

@@

We again see that the primary predictor in the logistic regression is the `nox` 
with `rad` and `ptratio` rounding out the top three predictors staying in line 
with the base model. Living close to the center of town, having teachers, and 
not having polluted air are the largest predictors in terms of staying above median 
crime rates.

@@

Model 3 will be a binary logistic regression allowing for all variables and 
selected pair-wise interactions pruned using both forward and backward stepwise 
regression based on AICc to select the optimal parameters. The captured indices 
of the test set for models 1 & 2 are used to split the data for model 3 as well. 
To address the possibility of over-fitting, we performed five-fold cross-validation 
on the training set.  With the vanilla logistic regression model in R, `glm` with 
a binomial family and a logit link, there are no other hyper parameters available 
for tuning.  More sophisticated logistic regression implementation have these 
parameters for fine-tuning.

Considering *all* interactions would be foolhardy. Given the need for dummy variables, 
there will 19 predictors and \(_{19}C_2 = 171\) possible pairs, too much. Therefore, 
intentional selection of pairs is required.

Looking at the predictors, there are some, which logically may interact with each other, 
such as:   
* `indus`
	* The concentration of non-retail businesses in the area
* `nox`
    		* Nitrous Oxide is a common industrial pollutant.

There is a high correlation in the training set between the two. Allowing for its 
interaction may temper the result of their relationship without losing more information.

@@

It seems that there may be reason to consider an interaction between `chas` and
`nox`. Can that many industrial plants be on the river?

Similarly with `indus` and `lstat`. Higher-cost homes tend to be further away from the traffic
and pollution of industrial areas.

@@

The selected model has some interesting results. First, the predictor with the 
greatest effect is `nox`.  Second, the only model variable which serves to 
**decrease** the crime rate from the baseline is `rm`.

The interactions between `indus` and both `nox` and `lstat` are significant. 
As theorized, these interactions serve to temper their individual contributions to 
crime. Remember, a `1` indicates crime above the median, so positive coefficients 
indicate contributions to *increasing* the crime rate. A similar observation holds 
`rm`. While `dis` on its own did not decrease the model's AICc, its interaction 
to temper `rm` did help the model's deviance. No other selected interactions are 
left in the model.

A number of variables fell out of the model, such as `chas`, `dis`, and `zn` 
among others. The remaining predictors are all significant, at least at the 10% 
level, except for `rad.24`. However, from the last trace of the stepping routine 
shown below, removing it would ***drastically*** reduce the models performance.

Lastly, while `rad.1` is the base case, `rad.2` and `rad.6` have disappeared as 
well. This indicates that both `rad = 2` and `rad = 6` should be considered 
indistinguishable from `rad = 1`.

```
                    Df Deviance    AIC
<none>                   90.565 122.56
+ dis                1   89.191 123.19
+ zn                 1   89.558 123.56
+ tax                1   89.700 123.70
- ptratio            1   93.763 123.76
+ age                1   90.156 124.16
+ `nox:chasOnRiver`  1   90.224 124.22
- `rm:dis`           1   94.254 124.25
+ chas.OnRiver       1   90.273 124.27
+ rad.6              1   90.315 124.31
- lstat              1   94.364 124.36
+ `ptratio:medv`     1   90.494 124.49
+ rad.2              1   90.554 124.55
- rm                 1   95.001 125.00
- `indus:lstat`      1   95.633 125.63
- `indus:nox`        1   97.351 127.35
- indus              1   97.597 127.60
- medv               1   98.568 128.57
- rad.3              1  101.594 131.59
- rad.5              1  105.594 135.59
- rad.7              1  118.832 148.83
- nox                1  134.786 164.79
- rad.8              1  149.967 179.97
- rad.4              1  152.757 182.76
- rad.24             1  152.797 182.80
```

@@

## Performance 
Below we show the confusion matrices for the three models, followed by a table with
the selected performance.
